---
title: "Virus_data_analysis_V3"
author: "Theo"
date: "2025-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(pheatmap)
library(DESeq2)
library(vegan)     # For community ecology analysis (alpha/beta diversity)
library(ggpubr)    # For publication-ready plots
library(viridis)
library(RColorBrewer)
library(ggrepel)
library(colorspace)
#library(EnhancedVolcano)
```

```{r}
#######3## Data Pre-processing ##############
# 1. Load and clean data
kraken_data <- read_tsv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/braken_data_all_samples.tsv")

kraken_clean <- kraken_data %>%
  mutate(
    Species = ifelse(grepl("ricinus", Sample), "ricinus", "persulcatus"),
    Sex = ifelse(grepl("\\-f\\-", Sample), "Female", 
                ifelse(grepl("\\-m\\-", Sample), "Male", "Unknown")),
    Sample_ID = str_extract(Sample, "ric-[fm]-\\d+|pers-[fm]-\\d+")
  )

# 2. Check data structure
glimpse(kraken_clean)

# 3. Filter for species-level data and aggregate
species_data <- kraken_clean %>%
  filter(Rank == "S") %>%
  group_by(Sample_ID, Species, Sex, NCBI_TaxID, Scientific_Name) %>%
  summarise(Total_Reads = sum(Reads_Count), .groups = "drop")

# 4. Create the virus matrix - SLOW BUT RELIABLE METHOD
# First, get all unique samples and viruses
all_samples <- unique(species_data$Sample_ID)
all_viruses <- unique(species_data$Scientific_Name)

# Create an empty matrix
virus_matrix <- matrix(0, nrow = length(all_samples), ncol = length(all_viruses))
rownames(virus_matrix) <- all_samples
colnames(virus_matrix) <- all_viruses

# Fill the matrix manually
for(i in 1:nrow(species_data)) {
  sample <- species_data$Sample_ID[i]
  virus <- species_data$Scientific_Name[i]
  count <- species_data$Total_Reads[i]
  virus_matrix[sample, virus] <- count
}

# Convert to data frame and add Sample_ID as a column
virus_df <- as.data.frame(virus_matrix) %>%
  rownames_to_column("Sample_ID")

# 5. Save your results
write_csv(kraken_clean, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/1000_reads_braken_clean_data.csv")
write_csv(species_data, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/1000_reads_species_level_data.csv")
write_csv(virus_df, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/1000_read_virus_abundance_matrix_df.csv")

```




## Goal 1 ###############
#Panels
#(A) Stacked barplot: relative abundance of top 10 viral taxa across samples, grouped by species.
#(B) Barplot: top 10 most abundant viruses overall (mean relative abundance ± SD).
#Purpose: Show composition and which viruses dominate across samples/species.

#Inputs
#virus_matrix.csv (counts)
#metadata.tsv

#Method
#Convert counts → relative abundance per sample (percent)
#Collapse low-abundance taxa into "Other" (e.g. sum of taxa <1% mean)
#Order samples by species then by alpha diversity or sample ID

```{r}
# prep

mat <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/1000_read_virus_abundance_matrix_df.csv", row.names=1, check.names=FALSE)
meta <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/sample_metadata.csv")
# long format and relative abundance
long <- mat %>% rownames_to_column("Sample_ID") %>%
  pivot_longer(-Sample_ID, names_to="Virus", values_to="Reads") %>%
  left_join(meta, by="Sample_ID") %>%
  group_by(Sample_ID) %>%
  mutate(RelAb = 100 * Reads / sum(Reads)) %>% ungroup()
```

```{r}
# subplot A: pick top 10 by mean rel abundance 
# top10 viruses by mean relative abundance

# ---- Select top 10 viruses by mean relative abundance ----
top10 <- long %>%
  group_by(Virus) %>%
  summarise(meanRel = mean(RelAb, na.rm = TRUE)) %>%
  arrange(desc(meanRel)) %>%
  slice_head(n = 10) %>%
  pull(Virus)

# ---- Collapse others into "Other" and calculate total reads ----
long_plot <- long %>%
  mutate(Virus = if_else(Virus %in% top10, Virus, "Other")) %>%
  group_by(Sample_ID) %>%
  mutate(TotalReads = sum(RelAb, na.rm = TRUE)) %>%
  ungroup()

# ---- Set factor levels so "Other" is last ----
virus_levels <- c(sort(unique(top10)), "Other")
long_plot$Virus <- factor(long_plot$Virus, levels = virus_levels)

# ---- Generate colors ----
n_top <- length(top10)
top_colors <- colorRampPalette(brewer.pal(8, "Paired"))(n_top)
names(top_colors) <- top10

fill_colors <- c(top_colors, Other = "grey70")

# ---- Plot ----
p <- ggplot(long_plot, aes(x = fct_reorder(Sample_ID, TotalReads),
                           y = RelAb,
                           fill = Virus)) +
  geom_col() +
  facet_grid(~Species, scales = "free_x", space = "free") +
  theme_minimal() +
  labs(y = "Relative abundance (%)",
       x = "Sample (ordered by total abundance)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1)) +
  scale_fill_manual(values = fill_colors)

p

ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/top10_viruses_by_mean_relative_abundance.pdf",
       plot = p,
       width = 12,
       height = 6,
       dpi = 300)
```


```{r}
###############################################################################

# subplot B: barplot mean relative abundance
summary <- long %>% group_by(Virus) %>% summarise(meanRel=mean(RelAb), sdRel=sd(RelAb)) %>%
  arrange(desc(meanRel)) %>% slice_head(n=10)
ggplot(summary, aes(x=reorder(Virus, meanRel), y=meanRel)) +
  geom_col() + geom_errorbar(aes(ymin=meanRel-sdRel, ymax=meanRel+sdRel), width=.2) +
  coord_flip() + theme_minimal() + labs(y="Mean relative abundance (%)")
```
```{r}
ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/mean_relative_abundance_updated2.pdf",
        width = 12,       # width in inches
       height = 7,       # height in inches
       dpi = 300)        # resolution in dots per inch
```

## Goal 2 ##########

#########################################################################################################
#Figure 2 — Prevalence vs mean abundance (bubble plot)
#Panel
#Bubble chart where x = prevalence (fraction of samples positive), y = mean relative abundance, bubble size = mean reads or mean abundance, color = viral family.
#Purpose: Identify viruses that are both common (prevalent) and abundant.
#Inputs
#virus_matrix.csv
#metadata.tsv
#optional: virus_families.csv mapping scientific name → family
#Method
#Convert wide → long; compute per-virus prevalence and mean rel abundance
#Prevalence = (# of samples with Reads > 0) / total samples
#Optionally filter viruses with prevalence < 0.01 to declutter

```{r pressure, echo=FALSE}
long <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/1000_read_virus_abundance_matrix_df.csv", row.names=1, check.names=FALSE) %>%
  rownames_to_column("Sample_ID") %>%
  pivot_longer(-Sample_ID, names_to="Virus", values_to="Reads") %>%
  left_join(read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/sample_metadata.csv"), by="Sample_ID")

# numeric cleanup
long$Reads <- as.numeric(long$Reads); long$Reads[is.na(long$Reads)] <- 0
long <- long %>% group_by(Sample_ID) %>% mutate(RelAb=100*Reads/sum(Reads)) %>% ungroup()

summary <- long %>% group_by(Virus) %>%
  summarise(MeanRel = mean(RelAb), Prevalence = mean(Reads > 0), MeanReads = mean(Reads)) %>%
  left_join(read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/viral_families.csv"), by=c("Virus"="Scientific_Name"))

# bubble plot
ggplot(summary, aes(x = Prevalence,
                    y = MeanRel,
                    size = MeanReads,
                    color = Virus_Family)) +

  geom_point(alpha = 0.8) +

  geom_text_repel(aes(label = ifelse(Prevalence > 0.1 | MeanRel > 1, Virus, "")),
                  size = 3,
                  max.overlaps = Inf) +

  scale_size(range = c(3, 12)) +

  scale_color_viridis_d(option = "turbo") +   # many distinct colors

  theme_minimal() +

  labs(x = "Prevalence",
       y = "Mean relative abundance (%)",
       color = "Virus family",
       size = "Mean reads")
```

```{r}

```


```{r}
# Keep only viruses with non-zero mean relative abundance
summary_nonzero <- summary %>%
  filter(MeanRel > 0)

ggplot(summary_nonzero, aes(x = Prevalence,
                    y = MeanRel,
                    size = MeanReads,
                    color = Virus_Family)) +

  geom_point(alpha = 0.8) +

  geom_text_repel(aes(label = ifelse(Prevalence > 0.1 | MeanRel > 1, Virus, "")),
                  size = 3,
                  max.overlaps = Inf) +

  scale_size(range = c(3, 12)) +

  scale_color_viridis_d(option = "turbo") +   # many distinct colors

  theme_minimal() +

  labs(x = "Prevalence",
       y = "Mean relative abundance (%)",
       color = "Virus family",
       size = "Mean reads")
```

```{r}
ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/mean_relative_abundance_prevalence.pdf",
        width = 12,       # width in inches
       height = 7,       # height in inches
       dpi = 300)        # resolution in dots per inch
```

```{r}
# Filter out zeros and add pseudocounts for log scale
summary_nonzero <- summary %>%
  filter(MeanRel > 0, Prevalence > 0) %>%
  mutate(
    MeanRel = MeanRel + 1e-6,
    Prevalence = Prevalence + 1e-6
  )

# Generate as many distinct colors as families
n_families <- length(unique(summary_nonzero$Virus_Family))
family_palette <- qualitative_hcl(n_families, palette = "Dark3")

ggplot(summary_nonzero, aes(x = Prevalence, y = MeanRel,
                            size = MeanReads,
                            color = Virus_Family)) +
  geom_point(alpha = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(Prevalence > 0.1 | MeanRel > 1, Virus, "")),
    size = 3, max.overlaps = 50
  ) +
  scale_size(range = c(3, 12)) +
  scale_color_manual(values = family_palette) +
  scale_y_log10() +
  scale_x_log10() +
  theme_minimal(base_size = 14) +
  labs(x = "Prevalence [log10 scale]",
       y = "Mean relative abundance (%) [log10 scale]",
       color = "Virus Family",
       size = "Mean Reads") +
  guides(
    color = guide_legend(ncol = 1, override.aes = list(size = 3)),
    size = guide_legend(override.aes = list(color = "grey40"))
  )
```

```{r}
ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/relative_abundance_bubble_plot_byFamily.png",
        width = 14,       # width in inches
       height = 8,       # height in inches
       dpi = 300)        # resolution in dots per inch
```


## Goal 3 ############

##############################################################################################

#Figure 3 — Heatmap of viral abundance (samples × viruses)
#Panel
#Heatmap (log2(count+1)) with samples as rows and top viruses as columns; sample annotation bars for species, sex.
#Purpose: Visualize sample-level patterns and clusters.
#Inputs
#virus_matrix.csv
#metadata.tsv

#Method
#Select top N viruses by total abundance (e.g., top 50)
#Make numeric matrix, rownames = Sample_ID
#Row annotation: species, sex (from metadata)
#Scale rows or columns depending on interpretation (commonly scale rows to compare relative dominance per sample)
```{r}

mat <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/1000_read_virus_abundance_matrix_df.csv", row.names=1, check.names=FALSE)
meta <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/sample_metadata.csv") %>% column_to_rownames("Sample_ID")

# compute abundance per virus read
virus_totals <- colSums(mat, na.rm = TRUE)


#Select to 50 most abundant viral reads
topN <- names(sort(virus_totals, decreasing = TRUE))[1:50]

#for all 49 viruses:
#topN <- names(sort(virus_totals, decreasing = TRUE))


# keep only numeric columns, then convert to matrix
mat_top <- mat[, topN, drop = FALSE] %>%     # keep as data.frame if single column
  mutate(across(everything(), ~ as.numeric(.))) %>%
  as.matrix()

# optional: ensure rownames are preserved
rownames(mat_top) <- rownames(mat)[rownames(mat_top)]

# annotation
ann <- meta[, c("Species","Sex")] 

pheatmap(log2(mat_top+1), annotation_row=ann,
         scale="row", clustering_method="complete",
         filename="fig3_#heatmap.pdf", width=12, height=10)

```

##Heatmap modify
```{r}
library(pheatmap)
library(RColorBrewer)

mat_plot <- log2(mat_top + 1)
mat_plot[mat_plot > 10] <- 10   # clamp to 0–10

ann$Sex <- factor(ann$Sex, levels = c("male", "female"))
ann <- ann[rownames(mat_plot), , drop=FALSE]

ann_colors <- list(
  Sex = c(male = "#1f78b4", female = "#e31a1c")
)

## --- Define color palette with plain color at zero ---
breaks <- seq(0, 10, length.out = 101)
cols <- colorRampPalette(c("white", "yellow", "orange", "red", "darkred"))(100)

## --- Plot heatmap ---
pheatmap(log2(mat_top + 1),
         annotation_row = ann,
         annotation_colors = ann_colors,
         scale = "none",                  # important: do not rescale rows
         clustering_method = "complete",
         color = cols,
         breaks = breaks,
         filename = "fig3_1_heatmap.pdf",
         width = 12,
         height = 10)
```

# 4

###############################################################################################################


#Figure 4 — Differential abundance (species comparison)
#Panels
#Volcano plot of differential viruses (I. ricinus vs I. persulcatus)
#Table inset or top hits barplot

#Purpose:
#Identify viruses significantly enriched in one tick species.

#Inputs:
#Raw counts matrix (no prior normalization)
#metadata.tsv with Species factor

#Method:
#Use DESeq2 on counts matrix (or edgeR)
#Filter low-count viruses (e.g., keep viruses with counts >= 10 in at least 3 samples)
#Design formula: ~ Species (or ~ Species + Sex + Batch if you have covariates)
#Adjusted p-value threshold: FDR < 0.05

```{r}

cts <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/1000_read_virus_abundance_matrix_df.csv", row.names=1, check.names=FALSE)
meta <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/sample_metadata.csv") %>% column_to_rownames("Sample_ID")

dim(cts)
dim(meta)

# transpose so that rows = viruses, cols = samples
cts_t <- t(cts)
dim(cts_t)
# set proper names
colnames(cts_t) <- rownames(meta)

# ensure alignment
stopifnot(all(colnames(cts_t) == rownames(meta)))

# construct DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = round(cts_t),
                              colData = meta,
                              design = ~ Species)   # you can add + Sex if you want to adjust

# -------------------------------------------------
# Pre-filtering
# -------------------------------------------------
# remove viruses with very few reads across all samples
dds <- estimateSizeFactors(dds, type = "poscounts")

dds <- dds[rowSums(counts(dds)) > 1000, ]

# -------------------------------------------------
# Normalization & QC
# -------------------------------------------------
dds <- DESeq(dds)

# Variance stabilizing transformation
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)

# PCA plot
plotPCA(vsd, intgroup = "Species") + ggtitle("PCA of Viral Communities")

# Sample distance heatmap
sampleDists <- dist(t(assay(vsd)))

sample_sample_dist_heatmap <- pheatmap(as.matrix(sampleDists),
         annotation_col = meta,
         main = "Sample-to-sample distances")

```

```{r}
pdf("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/sample_sample_dist_heatmap.pdf",
     width = 12, height = 8)
print(sample_sample_dist_heatmap)
dev.off()
```



```{r}

# -------------------------------------------------
# Differential abundance testing
# -------------------------------------------------
# Extract results for Species effect
res <- results(dds, contrast = c("Species", "ricinus", "persulcatus"))

# order by adjusted p-value
res <- res[order(res$padj), ]

# Save results
write.csv(as.data.frame(res), "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/DESeq2_results_species.csv")

# -------------------------------------------------
# Visualization
# -------------------------------------------------

# MA plot
plotMA(res, ylim = c(-4, 4))

# Volcano plot
EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = 1,
                title = "Virus abundance differences between tick species",
                subtitle = "DESeq2 results",
                pointSize = 2.0,
                labSize = 3.5)

# Heatmap of top 20 differentially abundant viruses
topVar <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 20)
Top_20_diff_heatmap <- pheatmap(assay(vsd)[topVar, ],
         annotation_col = meta,
         scale = "row",
         show_rownames = TRUE,
         clustering_method = "complete",
         main = "Top 20 variable viruses")

# -------------------------------------------------
# Export normalized counts
# -------------------------------------------------
norm_counts <- counts(dds, normalized=TRUE)
write.csv(as.data.frame(norm_counts), "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/Normalized_counts.csv")
```

```{r}
pdf("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/Top_20_diff_Heatmap.pdf",
    width = 12, height = 8)
print(Top_20_diff_heatmap)
dev.off()

```

#### Phase C: Advanced Statistical Analysis ##########

1. Alpha Diversity: Within-sample diversity
```{r}
virus_matrix <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/virus_matrix_data_1000.csv", row.names=1, check.names=FALSE)

meta <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/sample_metadata.csv") #%>% column_to_rownames("Sample_ID")
```

```{r}
# Keep only viral count columns
mat_clean <- virus_matrix %>%
  as.data.frame() %>%
  select(where(is.numeric))   # keep numeric columns only

# bring back sample IDs as a column
mat_clean <- mat_clean %>% rownames_to_column("Sample_ID")

alpha_diversity <- mat_clean %>%
  mutate(
    Shannon = diversity(select(., -Sample_ID), index = "shannon"),
    Simpson = diversity(select(., -Sample_ID), index = "simpson"),
    Richness = rowSums(select(., -Sample_ID) > 0)
  )

alpha_diversity <- alpha_diversity %>%
  left_join(
    meta %>% select(Sample_ID, Species, Sex) %>% distinct(),
    by = "Sample_ID"
  )

# Save the alpha diversity results
write_csv(alpha_diversity, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/alpha_diversity_results.csv")

# View the results
head(alpha_diversity)
```


```{r}
# Test for statistical differences
shannon_model <- aov(Shannon ~ Species * Sex, data = alpha_diversity)
summary(shannon_model)

# Or use a non-parametric test if assumptions are violated
#kruskal_wallace_test <- kruskal.test(Shannon ~ Species, data = alpha_diversity)
#summary(kruskal_wallace_test)
```
           Df Sum Sq Mean Sq F value Pr(>F)  
Species      1  0.852  0.8519   4.547 0.0357 *
Sex          1  0.179  0.1795   0.958 0.3304  
Species:Sex  1  0.356  0.3563   1.902 0.1713  
Residuals   89 16.675  0.1874                 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1



```{r}
# Plot the results
P1 <- ggplot(alpha_diversity, aes(x = Species, y = Shannon, fill = Sex)) +
  geom_boxplot() +
  stat_compare_means(method = "kruskal.test", label = "p.format") +
  labs(title = "Alpha Diversity by Species & Sex") +
  theme_bw()
print(P1)
```

```{r}
# save the plot
ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/Alpha_diversity_plot2.pdf",P1,
  width  = 8,
  height = 6,
  units  = "in"
)
```

#############################################################################
2. Beta Diversity: Between-sample diversity
```{r}
# Perform PERMANOVA (adonis) to test if communities differ by species/sex
# This uses Bray-Curtis dissimilarity
# 1. Make sure Sample_ID is preserved
mat_clean <- virus_matrix %>%
  as.data.frame() %>%
  rownames_to_column("Sample_ID")   # If IDs are in rownames

# 2. Drop metadata (Species, Sex) if they are inside virus_matrix
mat_clean <- mat_clean %>%
  select(-matches("Species|Sex"))   # remove non-numeric metadata columns



# Verify the dimensions match
cat("After correction - Community matrix samples:", nrow(mat_clean), "\n")
cat("After correction - Metadata samples:", nrow(meta), "\n")

# They should now be equal
```

```{r}
# PERMANOVA test
# Ensure mat_clean has only numeric viral counts
mat_numeric <- mat_clean %>%
  select(-Sample_ID) %>%
  as.matrix()

# Match rownames to Sample_ID
rownames(mat_numeric) <- mat_clean$Sample_ID

# Make sure meta rows align
meta <- meta %>%
  filter(Sample_ID %in% rownames(mat_numeric)) %>%
  arrange(match(Sample_ID, rownames(mat_numeric)))

# Double-check alignment
stopifnot(all(rownames(mat_numeric) == meta$Sample_ID))

# Run PERMANOVA (Bray–Curtis distance)
permanova_result <- adonis2(
  mat_numeric ~ Species * Sex,
  data = meta,
  method = "bray",
  permutations = 999
)

# Run PERMANOVA (Bray–Curtis distance)
permanova_result_sex <- adonis2(
  mat_numeric ~ Sex,
  data = meta,
  method = "bray",
  permutations = 999)

print(permanova_result_sex)

# Run PERMANOVA (Bray–Curtis distance)
permanova_result_Species <- adonis2(
  mat_numeric ~ Species,
  data = meta,
  method = "bray",
  permutations = 999)

print(permanova_result_Species)

# Save PERMANOVA results
permanova_df <- as.data.frame(permanova_result) %>%
  rownames_to_column("Term")
write_csv(permanova_df, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/viral_analysis/permanova_results.csv")
```

```{r}
library(dplyr)
library(tibble)
# NMDS ordination
# NMDS ordination - CORRECTED
set.seed(123) # For reproducible results

# Ensure community_matrix is numeric
mat_numeric <- mat_clean %>%
  mutate(across(everything(), as.numeric))

# Create proper metadata
#metadata_corrected <- kraken_clean %>%
 # select(Sample_ID, Species, Sex) %>%
  #distinct() %>%
  #filter(Sample_ID %in% rownames(community_matrix_numeric)) %>%
  #column_to_rownames("Sample_ID")

mat_numeric <- mat_clean %>%
  column_to_rownames("Sample_ID") %>%  # only if Sample_ID is a column
  t() %>%
  as.data.frame()

# Now samples should be rows
mat_numeric <- as.matrix(cts)

# Check dimensions
dim(mat_numeric)   # should be 92 x 49 (samples x viruses)


# Align metadata
meta <- meta %>%
  filter(Sample_ID %in% rownames(mat_numeric)) %>%
  arrange(match(Sample_ID, rownames(mat_numeric)))

dim(meta)
# Run NMDS
# Bray-Curtis
dist_mat <- vegdist(mat_numeric, method = "bray")
nmds <- metaMDS(dist_mat, k = 2, trymax = 100)

nmds_df <- scores(nmds) %>% as.data.frame()

if (!"Sample_ID" %in% colnames(nmds_df)) {
  nmds_df <- rownames_to_column(nmds_df, "Sample_ID")
}

meta_df <- meta

if (!"Sample_ID" %in% colnames(meta_df)) {
  meta_df <- rownames_to_column(meta_df, "Sample_ID")
}

nmds_scores <- left_join(nmds_df, meta_df, by = "Sample_ID")

nrow(scores(nmds))
# Extract NMDS scores (coordinates)



# Save the results
write_csv(nmds_scores, "/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/nmds_scores.csv")

# Check if it worked
cat("NMDS completed successfully!\n")
cat("Stress value:", nmds$stress, "\n")
print(head(nmds_scores))
```

```{r}
nmds_scores <- read.csv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/nmds_scores.csv")

# NMDS plot
p2 <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = Sex, shape = Species)) +
  geom_point(size = 3) +
  stat_ellipse() +
  theme_bw() +
  labs(title = "NMDS of Viral Communities") # subtitle = paste("Stress:", round(nmds$stress, 3))
print(p2)
# Save plots
ggsave("alpha_diversity_plot.png", P1, width = 8, height = 6)
ggsave("nmds_plot.png", p2, width = 8, height = 6)

#library(gridExtra)
#combined_plot <- grid.arrange(P1, p2, ncol = 2)
ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/nmds_combined2.pdf", p2, width = 10, height = 6, dpi = 300)
message("Analysis complete! All results saved to CSV files.")
```


# Unmapped fastq_metrics
```{r}
df <- read_tsv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/hisat2_ricinus_mapping_stats_summary.tsv")

# Basic stats
summary(df$Overall_Alignment_Rate)
mean(df$Unmapped_Reads)
# Plot
ggplot(df, aes(x=reorder(Sample, Overall_Alignment_Rate), y=Overall_Alignment_Rate)) +
  geom_col(fill="darkcyan") +
  coord_flip() +
  labs(y="Overall Alignment Rate (%)", x="Sample") +
  theme_minimal(base_size=12)

sum(df$Unmapped_Reads)
summary(df$Unmapped_Reads)
mean(df$Unmapped_Reads)

```
```{r}
# Load packages
library(tidyverse)
library(ggthemes)

# Read HISAT2 summary file
stats <- read_tsv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/hisat2_persul_mapping_summary.tsv")

# Convert numeric columns
stats <- stats %>%
  mutate(across(c(Total_Reads, Aligned_Concordant_0, Aligned_Concordant_1,
                  Aligned_Concordant_Multi, Aligned_Discordant_1,
                  Unmapped_Reads, Overall_Alignment_Rate), as.numeric))

# Calculate total mapped reads
stats <- stats %>%
  mutate(Mapped_Reads = Aligned_Concordant_1 + Aligned_Concordant_Multi + Aligned_Discordant_1)

# Add species info from sample names
stats <- stats %>%
  mutate(Species = case_when(
    grepl("pers", Sample, ignore.case = TRUE) ~ "I. persulcatus",
    grepl("ric", Sample, ignore.case = TRUE) ~ "I. ricinus",
    TRUE ~ "Unknown"
  ))

# ====== 1️⃣ Absolute counts (in millions) ======
stats_long_counts <- stats %>%
  select(Sample, Species, Mapped_Reads, Unmapped_Reads) %>%
  pivot_longer(cols = c(Mapped_Reads, Unmapped_Reads),
               names_to = "Category", values_to = "Reads")

p1 <- ggplot(stats_long_counts, aes(x = reorder(Sample, -Reads), y = Reads / 1e6, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~Species, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("Mapped_Reads" = "#0072B2", "Unmapped_Reads" = "#D55E00"),
                    name = "Read Category",
                    labels = c("tick Reads", "non-tick Reads")) +
  labs(
    title = "Persulcatus tick reads vs non-tick Reads per Sample",
    x = "Sample ID",
    y = "Read Count (Millions)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.grid.major.x = element_blank())

ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/mapped_vs_unmapped_reads_persulcatus.pdf", p1, width = 12, height = 6, dpi = 300)


# ====== 2️⃣ Percent stacked bar plot ======
stats_long_perc <- stats %>%
  mutate(Mapped_Percent = (Mapped_Reads / Total_Reads) * 100,
         Unmapped_Percent = (Unmapped_Reads / Total_Reads) * 100) %>%
  select(Sample, Species, Mapped_Percent, Unmapped_Percent) %>%
  pivot_longer(cols = c(Mapped_Percent, Unmapped_Percent),
               names_to = "Category", values_to = "Percent")

p2 <- ggplot(stats_long_perc, aes(x = reorder(Sample, -Percent), y = Percent, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~Species, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("Mapped_Percent" = "#1f78b4", "Unmapped_Percent" = "#e31a1c"),
                    name = "Read Category",
                    labels = c("tick (%)", "non-tick (%)")) +
  geom_text(aes(label = sprintf("%.1f%%", Percent)),
            position = position_stack(vjust = 0.5),
            color = "white", size = 2.8) +
  labs(
    title = "Percentage of persulcatus tick vs non-tick Reads per Sample",
    x = "Sample ID",
    y = "Percentage (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.grid.major.x = element_blank())

ggsave("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/viral_analysis/mapped_vs_unmapped_reads_persulcatus_percent.pdf", p2, width = 12, height = 6, dpi = 300)

# Print plots in interactive session
p1
p2

```

#

##################################################################################################
#Figure 5 — Novel virus discovery and phylogeny

#1. Assemble Viral Reads
#After extracting viral reads from Kraken2 or host-filtered datasets:
#Tools: metaSPAdes (for metatranscriptomic/viral assembly)
```{bash}
#Assemble viral reads
bash script


#1. NCBI RefSeq Viral Protein Database (Recommended for DIAMOND/BLASTx)
mkdir database
cd database
wget ftp://ftp.ncbi.nlm.nih.gov/refseq/release/viral/viral.1.protein.faa.gz

gunzip viral.*.faa.gz
cat viral.*.faa > viral_refseq.faa

module load diamond
diamond makedb --in viral_refseq.faa -d viral_refseq.dmnd

```



--------------------------------------------------------------------
#2 Quality Filter and Remove Short/Low-Complexity Contigs
Remove assemblies shorter than 300–500 bp
```{bash}
seqtk seq -L 500 contigs.fasta > contigs.500_plus.filt_persul.fasta
```

#3. Annotate Contigs to Identify Viral Genes
Use Diamond blastx against viral protein databases (RefSeq Viral)

```{bash}
diamond blastx -d viral.dmnd -q contigs.500_plus.filt_persul.fasta \
 -o contigs_viral.m8 -f 6 qseqid sseqid pident length evalue bitscore stitle \
 --max-target-seqs 20 --evalue 1e-5 --threads 16
 
# Test hit identity values/coverage percentage 
 awk '{print $3}' contigs_viral.m8 | sort -n | uniq | head
```

#4. Extract RdRp Gene (or Other Marker Genes)
The RNA-dependent RNA polymerase (RdRp) is the universal marker gene for RNA viruses.
```{bash}
#Search annotation file (m8) for RdRp-like hits
#Extract matching contigs
#Translate nucleotide contigs → protein sequences

awk '$3 >= 80 && $5 == 0.0' contigs_viral.m8 | grep -i -E "rdrp|rna-dependent|polymerase" | cut -f1 |sort -u > persul_nairo_hit_rdrp_ids.txt

#grep -i -E "rdrp|rna-dependent|polymerase" contigs_viral.m8 | awk '$3 >= 90 && #$11 == 0.00' | cut -f1 | sort -u > persul_nairo_rdrp_ids.txt

seqtk subseq contigs.500_plus.filt_persul.fasta persul_nairo_hit_rdrp_ids.txt > persul_beiji_rdrp_contigs.fasta

```



# Finding ORFs using EMBOSS getorf
# Extracting amino acid sequences
# Filtering for RdRp or viral proteins
# Preparing sequences for alignment and phylogeny

```{bash}
#5. Load EMBOSS on Puhti

module load emboss

#. Run getorf to find ORFs
#You have assembled contigs (contigs.fasta).
#You now extract all possible ORFs ≥300 nt (or ≥600 nt for viruses if needed).

getorf -sequence combined_persul_contigs.fasta -outseq persul_viral_ORFs.faa -minsize 900
  
seqkit stats persul_viral_ORFs.faa

#-minsize 900 → keeps ORFs ≥900 bp (≈100 aa), appropriate for RdRp detection
#-find 3 → finds ORFs on all six frames
#Output gives amino acid sequences directly (no nucleotide sequences)


# Sanitize FASTA headers (Optional but recommended)
sed -n 's/^>\([^ ]*\).* \([A-Za-z0-9_. -]*virus\).*/>\1_\2/p' full_reference_nt.fasta | sed 's/[ ()]/_/gI' > reference_short.fasta

sed -n 's/.*protein_id=\([^]]*\).*/>\1_RdRp/p' nairoviridae_ref_prot.faa > nairoviridae_ref_prot_short.faa

#This removes spaces and keeps only the ORF ID.
```


```{bash}

#6. Identify RdRp or viral proteins using DIAMOND
#Now annotate all ORFs and select only those matching viral proteins.

diamond blastp \
  -q un_nairo_contigs_orfs_clean.faa \
  -d /scratch/project_2005863/metatranscriptomics_2025/Theo_analysis_metatranscriptomics/database/viral_refseq.dmnd \
  -o un_nairo_orf_viral_hits.m8 \
  -f 6 qseqid sseqid pident length evalue bitscore stitle \
  --max-target-seqs 10 --evalue 1e-5 --threads 16


# . Extract RdRp-positive ORFs
grep -i -E "rdrp|rna-dependent|polymerase" un_nairo_orf_viral_hits.m8 | cut -f1 | sort -u > orf_rdrp_ids.txt

#Extract only those ORFs:
seqtk subseq un_nairo_contigs_orfs_clean.faa orf_rdrp_ids.txt > nairo_uncl_rdrp_orfs.faa

#. Remove duplicates and very short sequences
seqkit rmdup nairo_uncl_rdrp_orfs.faa > rdrp_orfs.unique.faa

#Remove sequences <100 aa:
seqkit seq -m 100 rdrp_orfs.unique.faa > rdrp_orfs.filtered.faa

#Combine with reference RdRp sequences
cat rdrp_orfs.filtered.faa ref_rdrp.faa > rdrp_plus_refs.faa
```

```{bash}
#7. Align the protein sequences With MAFFT:
mafft --auto --thread 8 rdrp_plus_refs.faa > rdrp_aligned.faa

#Running alignment for multiple files
for unaligned in *_Rdrp.faa; do mafft --maxiterate 1000 --localpair --thread 8 --leavegappyregion $unaligned > ${unaligned%_Rdrp.faa}_aligned.faa; done

#Trim poorly aligned regions Using trimAl:
trimal -automated1 \
  -in rdrp_aligned.faa \
  -out rdrp_aligned_trimmed.faa
#Trimming with clipkit
module load clipkit
  clipkit Nairoviridae_aligned.faa -m gappy -g 0.5 -o Nairoviridae_aligned_trimmed.faa
```

```{bash}
#10. Build phylogenetic tree Using IQ-TREE 2:
module load iqtree
iqtree2 -s rdrp_aligned_trimmed.faa \
  -m MFP -bb 1000 -alrt 1000 -nt AUTO \
  -pre rdrp_tree

```

