---
title: "Metatranscriptomics_analysis"
author: "Theo"
date: "2025-08-28"
output: html_document
---
#### Phase A: Import and inspect data ##########

1. Load and Clean the Data in R
```{r}
# Load necessary libraries
library(tidyverse) # For data manipulation and plotting
library(vegan)     # For community ecology analysis (alpha/beta diversity)
library(ggpubr)    # For publication-ready plots
library(viridis)# For colorblind-friendly color scales
#library(heatmap)
```

2. Import the data
```{r}
list.files()
# Read the CSV file
kraken_data <- read_tsv("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/braken_data_all_samples.tsv")
problems(kraken_data)
head(kraken_data)
```

3. Re-form the data
```{r}
# Clean the Sample names: Extract meaningful identifiers
# This creates new columns for species (ricinus/persulcatus) and sex (f/m)
kraken_clean <- kraken_data %>%
  mutate(
    # Extract species: "ricinus" or "persulcatus"
    Species = ifelse(grepl("ricinus", Sample), "ricinus", "persulcatus"),
    # Extract sex: "F" or "M" (assuming 'f' and 'm' in filename)
    Sex = ifelse(grepl("\\-f\\-", Sample), "Female", 
                ifelse(grepl("\\-m\\-", Sample), "Male", "Unknown")),
    # Extract a simple sample ID (e.g., ric-f-10)
    Sample_ID = str_extract(Sample, "ric-[fm]-\\d+|pers-[fm]-\\d+")
  )

# View the structure of your data
glimpse(kraken_clean)
head(kraken_clean)
```


4. Filter and Aggregate Data
```{r}
# Data 1: Analyze at the Species level (most common)
species_data_1000 <- kraken_clean %>%
  filter(Rank == "S") %>% # Keep only rows with Species-level assignments
  group_by(Sample_ID, Species, Sex, NCBI_TaxID, Scientific_Name) %>%
  summarise(Total_Reads = sum(Reads_Count), .groups = "drop")

head(species_data_1000)

write_csv(species_data_1000, "species_level_data_1000.csv")

# Data 2: Create a prevalence table (Samples as rows, Viruses as columns)
# This is essential for most downstream analyses
virus_matrix_1000 <- species_data_1000 %>%
  ungroup() %>%
  select(Sample_ID, Species, Sex, Scientific_Name, Total_Reads) %>%
  pivot_wider(
    names_from = Scientific_Name, 
    values_from = Total_Reads,
    values_fn = list(Total_Reads = sum),  # Specify aggregation function
    values_fill = 0
  )

# The virus_matrix is now ready for diversity and differential abundance analysis.
head(virus_matrix_1000)

write_csv(virus_matrix_1000, "virus_matrix_data_1000.csv")
```

#### Phase B: Exploratory Data Analysis (EDA) ##########

5. Basic Summary Statistics
```{r}
# Total viral reads per sample
total_reads_per_sample <- 
  species_data_1000 %>%
  group_by(Sample_ID, Species, Sex) %>%
  summarise(Total_Reads = sum(Total_Reads), .groups = "drop")
head(total_reads_per_sample)
```

```{r}
# Number of different virus species per sample (Richness)
richness_per_sample <- species_data_1000 %>%
  group_by(Sample_ID, Species, Sex) %>%
  summarise(Richness = n_distinct(Scientific_Name), .groups = "drop")


```

```{r}
# Combine and plot
summary_df <- total_reads_per_sample %>%
  left_join(richness_per_sample, by = c("Sample_ID", "Species", "Sex"))

readsPerSample <-
  ggplot(summary_df, aes(x = Species, y = Total_Reads, fill = Sex)) +
  geom_boxplot() +
  scale_y_log10() + # Often needed due to large range in read counts
  labs(title = "Total Viral Load by Tick Species and Sex",
       y = "Total Reads (log10 scale)") +
  theme_bw()

RichnessPerSample <-
  ggplot(summary_df, aes(x = Species, y = Richness, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Viral Richness by Tick Species and Sex",
       y = "Number of Virus Species") +
  theme_bw()

Plot1 <- ggpubr::ggarrange(readsPerSample, RichnessPerSample)

# save the plot
ggsave(plot = Plot1, "Viral_richness_byTick_plot_1000_reads.pdf",
  width  = 10,
  height = 5,
  units  = "in"
)
```
These plots will tell you if one tick species or sex has a higher viral load or carries more viruses



6. Compositional Analysis: What viruses are there?
```{r}
# Calculate relative abundance of each virus in each sample
relative_abundance <- species_data_1000 %>%
  group_by(Sample_ID) %>%
  mutate(Rel_Abund = (Total_Reads / sum(Total_Reads)) * 100) %>%
  ungroup()


```

# This formula is wrong! Prevalence is wrong.
```{r}
# Find the most abundant viruses overall
#library(tidyverse)

# Summarize abundance & prevalence
top_viruses_1000 <- relative_abundance %>%
  group_by(Scientific_Name) %>%
  summarise(
    Mean_Abund = mean(Rel_Abund), 
    Prevalence = n_distinct(Sample_ID[Total_Reads > 0]) / n_distinct(Sample_ID)
  ) %>%
  filter(Mean_Abund > 0.1 | Prevalence > 0.1) %>%
  arrange(-Mean_Abund)
```

```{r}
# Bubble plot
ggplot(top_viruses_1000, aes(x = Prevalence, y = Mean_Abund, 
                        size = Mean_Abund, color = Prevalence)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = Scientific_Name), hjust = 1, vjust = 1, size = 2) +
  scale_size(range = c(2, 7)) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Virus Abundance vs Prevalence in Tick Metatranscriptomes",
    x = "Prevalence (proportion of samples)",
    y = "Mean Relative Abundance (%)",
    size = "Mean Abundance",
    color = "Virus Type"
  )
```
```{r}
virus_families <- data.frame(
  Scientific_Name = c("Beiji nairovirus", "Fairhair virus", "Grotenhout virus", 
                      "Ixovirus", "Ixovirus norvegiae", "Mikuni nairovirus",
                      "Norwavirus", "Norwavirus grotenhoutense", "Norwavirus sp.",
                      "Norway nairovirus 1", "Norway partiti-like virus 1",
                      "Orthonairovirus", "Phlebovirus", "Pustyn virus",
                      "Sara tick phlebovirus", "Sopafab virus", 
                      "Yichun nairovirus", "unclassified Nairovirus",
                      "unclassified Norwavirus", "unclassified Phlebovirus"),
  Virus_Family = c("Nairovirus", "Nairovirus", "Norwavirus", 
                   "Ixovirus", "Ixovirus", "Nairovirus",
                   "Norwavirus", "Norwavirus", "Norwavirus",
                   "Nairovirus", "Partiti-like", "Nairovirus",
                   "Phlebovirus", "Phlebovirus", "Phlebovirus",
                   "Sopafab", "Nairovirus", "Nairovirus",
                   "Norwavirus", "Phlebovirus"))
```

```{r}
top_viruses <- top_viruses_1000 %>%
  left_join(virus_families, by = "Scientific_Name")
```

```{r}
ggplot(top_viruses, aes(x = Scientific_Name, 
                        y = Prevalence, 
                        size = Mean_Abund, 
                        color = Virus_Family)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(3, 15)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  labs(title = "Virus Prevalence and Abundance by Family",
       x = "Virus",
       y = "Prevalence",
       size = "Mean Abundance (%)",
       color = "Virus Family")
```
###########################################################################
### Computing abundance & prevalence #################################
```{r}
library(pheatmap)

# Load data
virus_data <- read.delim("braken_data_all_samples.tsv", sep="\t", header=TRUE)

# ---- Abundance (relative abundance per sample) ----
virus_abundance <- virus_data %>%
  group_by(Sample_ID) %>%
  mutate(Rel_Abundance = (Total_Reads / sum(Total_Reads)) * 100) %>%
  ungroup()

# ---- Prevalence (fraction of samples where detected) ----
virus_prevalence <- virus_abundance %>%
  group_by(Scientific_Name) %>%
  summarise(
    Mean_Abundance = mean(Rel_Abundance),
    Prevalence = sum(Total_Reads > 0) / n_distinct(virus_abundance$Sample_ID)
  )

# ---- Save summary table ----
write.table(virus_prevalence, "virus_abundance_prevalence_summary.tsv",
            sep="\t", row.names=FALSE, quote=FALSE)

# ---- Heatmap of relative abundance ----
virus_matrix <- virus_abundance %>%
  select(Sample_ID, Scientific_Name, Rel_Abundance) %>%
  spread(Scientific_Name, Rel_Abundance, fill=0)

rownames_mat <- virus_matrix$Sample_ID
virus_matrix <- as.matrix(virus_matrix[,-1])
rownames(virus_matrix) <- rownames_mat

pheatmap(log10(virus_matrix+1),
         cluster_rows=TRUE, cluster_cols=TRUE,
         show_rownames=TRUE, show_colnames=TRUE,
         fontsize=8,
         main="Viral Abundance Heatmap (log10 scale)")

# ---- Bubble plot (Prevalence vs Abundance) ----
ggplot(virus_prevalence, aes(x=Scientific_Name, 
                             y=Prevalence, 
                             size=Mean_Abundance, 
                             color=Scientific_Name)) +
  geom_point(alpha=0.7) +
  scale_size(range=c(3,12)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size=8)) +
  labs(title="Virus Prevalence and Abundance",
       x="Virus", y="Prevalence",
       size="Mean Abundance (%)",
       color="Virus")

```



```{r}
# Plot composition for top viruses
plot_data <- relative_abundance %>%
  filter(Scientific_Name %in% top_viruses$Scientific_Name)

ggplot(plot_data, aes(x = Sample_ID, y = Rel_Abund, fill = Scientific_Name)) +
  geom_col() +
  facet_grid(~ Species + Sex, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Viral Community Composition",
       y = "Relative Abundance (%)", x = "Sample") 
# +
#  scale_fill_viridis(discrete = TRUE) # Use a better color scale

# save the plot
ggsave("Viral_community_composition_plot.pdf")
```

#### Phase C: Advanced Statistical Analysis ##########

1. Alpha Diversity: Within-sample diversity
```{r}
# Calculate diversity indices correctly
alpha_diversity <-
  virus_matrix %>%
  as_data_frame() %>%
  select(-Species, -Sex) %>%
  column_to_rownames("Sample_ID") %>%  # Make Sample_ID the row names
  mutate(across(everything(), as.numeric)) %>%  # Ensure all values are numeric
  # Calculate diversity metrics
  mutate(
    Shannon = diversity(., index = "shannon"),
    Simpson = diversity(., index = "simpson"),
    Richness = rowSums(. > 0)  # Number of non-zero species
  ) %>%
  rownames_to_column("Sample_ID") %>%  # Bring Sample_ID back as a column
  select(Sample_ID, Shannon, Simpson, Richness) %>%
  # Add metadata
  left_join(
    kraken_clean %>% select(Sample_ID, Species, Sex) %>% distinct(),
    by = "Sample_ID"
  )

# Save the alpha diversity results
write_csv(alpha_diversity, "alpha_diversity_results.csv")

# View the results
head(alpha_diversity)
```


```{r}
# Test for statistical differences
shannon_model <- aov(Shannon ~ Species * Sex, data = alpha_diversity)
summary(shannon_model)
# Or use a non-parametric test if assumptions are violated

# kruskal.test(Shannon ~ Species, data = alpha_diversity)
```
           Df Sum Sq Mean Sq F value Pr(>F)  
Species      1  0.852  0.8519   4.547 0.0357 *
Sex          1  0.179  0.1795   0.958 0.3304  
Species:Sex  1  0.356  0.3563   1.902 0.1713  
Residuals   89 16.675  0.1874                 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1



```{r}
# Plot the results
ggplot(alpha_diversity, aes(x = Species, y = Shannon, fill = Sex)) +
  geom_boxplot() +
  stat_compare_means(method = "kruskal.test", label = "p.format") +
  labs(title = "Alpha Diversity (Shannon Index) by Species and Sex") +
  theme_bw()
```
```{r}
# save the plot
ggsave(
  here("/scratch/project_2005863/metatranscriptomics_2025/Theo_analysis/Alpha_diversity_by_species_sex_plot.pdf"
  ),
  width  = 7,
  height = 5,
  units  = "in"
)
```

#############################################################################
2. Beta Diversity: Between-sample diversity
```{r}
# Perform PERMANOVA (adonis) to test if communities differ by species/sex
# This uses Bray-Curtis dissimilarity
community_matrix <- virus_df %>%
  column_to_rownames("Sample_ID") %>%
  mutate(across(everything(), as.numeric))
```

```{r}
metadata <- kraken_clean %>%
  select(Sample_ID, Species, Sex) %>%
  distinct() %>%  # Ensure no duplicates
  filter(Sample_ID %in% rownames(community_matrix)) %>%  # Only keep samples that are in community matrix
  column_to_rownames("Sample_ID")

# Verify the dimensions match
cat("After correction - Community matrix samples:", nrow(community_matrix), "\n")
cat("After correction - Metadata samples:", nrow(metadata), "\n")

# They should now be equal
```

```{r}
# PERMANOVA test
permanova_result <- adonis2(community_matrix ~ Species * Sex, 
                           data = metadata, 
                           method = "bray",
                           permutations = 999)

# Save PERMANOVA results
permanova_df <- as.data.frame(permanova_result) %>%
  rownames_to_column("Term")
write_csv(permanova_df, "permanova_results.csv")
```

```{r}
# NMDS ordination
# NMDS ordination - CORRECTED
set.seed(123) # For reproducible results

# Ensure community_matrix is numeric
community_matrix_numeric <- community_matrix %>%
  mutate(across(everything(), as.numeric))

# Create proper metadata
metadata_corrected <- kraken_clean %>%
  select(Sample_ID, Species, Sex) %>%
  distinct() %>%
  filter(Sample_ID %in% rownames(community_matrix_numeric)) %>%
  column_to_rownames("Sample_ID")

# Run NMDS
nmds <- metaMDS(community_matrix_numeric, distance = "bray", k = 2, trymax = 100)

# Get scores and join with CORRECTED metadata
nmds_scores <- scores(nmds) %>% 
  as.data.frame() %>% 
  rownames_to_column("Sample_ID") %>% 
  left_join(
    metadata_corrected %>% rownames_to_column("Sample_ID"), 
    by = "Sample_ID"
  )

# Save the results
write_csv(nmds_scores, "nmds_scores_corrected.csv")

# Check if it worked
cat("NMDS completed successfully!\n")
cat("Stress value:", nmds$stress, "\n")
print(head(nmds_scores))
```

```{r}
# 3. Save ALL your results
results_list <- list(
  kraken_clean = kraken_clean,
  species_data = species_data,
  virus_abundance = virus_df,
  alpha_diversity = alpha_diversity,
  permanova = permanova_df,
  nmds = nmds_scores
)

# Save each result to CSV
for(obj_name in names(results_list)) {
  file_name <- paste0(obj_name, ".csv")
  write_csv(results_list[[obj_name]], file_name)
  message("Saved: ", file_name)
}

# 4. Create some basic plots and save them too
# Alpha diversity plot
p1 <- ggplot(alpha_diversity, aes(x = Species, y = Shannon, fill = Sex)) +
  geom_boxplot() +
  stat_compare_means(method = "kruskal.test", label = "p.format") +
  labs(title = "Alpha Diversity (Shannon Index)") +
  theme_bw()

# NMDS plot
p2 <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = Species, shape = Sex)) +
  geom_point(size = 3) +
  stat_ellipse() +
  theme_bw() +
  labs(title = "NMDS of Viral Communities", 
       subtitle = paste("Stress:", round(nmds$stress, 3)))

# Save plots
ggsave("alpha_diversity_plot.png", p1, width = 8, height = 6)
ggsave("nmds_plot.png", p2, width = 8, height = 6)

message("Analysis complete! All results saved to CSV files.")
```


```
This will show you if the overall viral community is different between I. ricinus and I. persulcatus.


#3. Differential Abundance Analysis
To find viruses that are significantly more abundant in one group than another, use DESeq2 (for raw counts) or ALDEx2 (for compositional data).

```{r}
# Example with DESeq2 (using raw counts from species_data)
library(DESeq2)

# Create a count matrix for DESeq2
count_matrix <- species_data %>%
  select(Sample_ID, Scientific_Name, Total_Reads) %>%
  pivot_wider(names_from = Scientific_Name, values_from = Total_Reads, values_fill = 0) %>%
  column_to_rownames("Sample_ID")

# Create a metadata dataframe
coldata <- alpha_diversity %>% select(Species, Sex) %>% rownames_to_column("Sample_ID") %>% column_to_rownames("Sample_ID")

# Ensure order is the same
count_matrix <- count_matrix[rownames(coldata), ]

# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = t(count_matrix),
                              colData = coldata,
                              design = ~ Species) # Test for effect of Species

# Run analysis
dds <- DESeq(dds)
res <- results(dds, contrast = c("Species", "I. ricinus", "I. persulcatus"))
res_df <- as.data.frame(res) %>% rownames_to_column("Virus") %>% arrange(padj)

# View top differentially abundant viruses
head(res_df %>% filter(padj < 0.05), 10)
```


Next Steps and Interpretation:
Biological Questions: Frame your analysis around questions:

"Does the virome of I. ricinus differ significantly from I. persulcatus?" (PERMANOVA, NMDS)

"Which specific viruses are biomarkers for each tick species?" (DESeq2)

"Is viral load or diversity affected by the sex of the tick?" (Alpha diversity plots, stats)

Focus on Key Viruses: Identify the most interesting viruses from your results (e.g., pathogens like Severe acute respiratory syndrome-related coronavirus, Norway nairovirus 1, or highly abundant novel viruses).

Validation: Consider validating key findings with PCR if these are important biological conclusions.
